Suplementary Material 1: Code to Reproduce Analysis and Plots

### 2023-03-23 S1_replicate_stats_and_plots.Rmd
draw more plots move code to functions

---
title: 'Suplementary Material 1: Code to Reproduce Analysis and Plots'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 2)
options(scipen = 9)
options(knitr.kable.NA="")
```

@@

## Data import

@@

```{r}
ee_tbl <- twinR::build_data_summary.table(data_births_monthly_EE)
ee_tbl_nl <- twinR::build_data_summary.table(data_births_monthly_EE_not_last)
```

\tiny
```{r}

  eetbls <- rbind(ee_tbl[-1,4:ncol(ee_tbl)-1],ee_tbl_nl[-1,4:ncol(ee_tbl_nl)-1])
eetbls$Subset <- c("All data", "No last birth")
knitr::kable(eetbls,
             caption = paste0("Data summary table of Estonian data",
                              " with and without last born child")) 
```
\normalsize

## Data import Original

@@

\tiny
```{r}
knitr::kable(twinR::build_data_summary.table(data_births_monthly_not_last)[-2],
             caption = "Data sumary table without last born child")
```
\normalsize

@@

```{r}
#' Fit Predictions
#'
#' This function fits a model using the given formula and dataset and 
#' computes predictions. The model is fit using the \code{\link[spaMM]{fitme}}
#' function from the \strong{spaMM} package.
#' 
#' @param dataset A data frame containing the data to be used for fitting the model.
#' @param formula A formula specifying the model to be fit.
#' @param predict Logical value indicating whether to do predictions. Default is TRUE.
#' @param nb_boot Number of bootstrap samples to use when computing predictions.
#'  Default is 1000.
#' @param predictionsDir Directory where precomputed predictions are stored.
#'  Default is "./data/predictions".
#' 
#' @return A list containing the fitted model object and a data frame with
#'  computed predictions.
fitPredictions <- function(dataset, formula,
                           predict=TRUE,
                           nb_boot=1000,
                           predictionsDir = "./data/predictions"){

  if(!dir.exists(predictionsDir)) dir.create(predictionsDir)

  args <- list(formula = stats::as.formula(formula),
             data = dataset, 
             family = stats::binomial(link = "logit"),
             method = "PQL/L")

  fit <-  twinR::fit_model_safely(timeout = Inf, .args = args)

  fitName <- deparse(substitute(dataset))
  #TODO maybe add formula also to the fitName like:
  #fitName <- paste0(form2str(stats::as.formula(formula)),fitName)

  predDataFname <- paste0(predictionsDir,"/",fitName ,"data_fig.rds")
  if(!file.exists(predDataFname) & predict){
    min_births <- min(dataset$births_total)
    max_births <- max(dataset$births_total)
    nd <- data.frame(births_total = min_births:max_births)
    data_fig <- compute_predictions(fit,
                                    newdata = nd,
                                    nb_boot = nb_boot)

    saveRDS(data_fig, predDataFname)

  } else {
    if(file.exists(predDataFname)){
      warning("Pre-computed predictions returned from file:\n", predDataFname,
            "\n If you want to re-run this time intensive step delete the file!")
    data_fig <- readRDS(predDataFname)
    } else {
      data_fig <- list(results = NULL)
    }

  }

  #garbage collecion after spaMM multi-core proccess
  gcstuff <- gc(verbose=FALSE)

  return(list(fit=fit, results=data_fig$results))
}
```

@@

```{r}
## Estonia - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_fit <- fitPredictions(dmm_EE, formula)

@@

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_fit <- fitPredictions(data_births_monthly_EE, formula, predict = F)

@@

```{r}
## TwinR - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_fit <- fitPredictions(dmm_orig, formula)

@@

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_fit <- fitPredictions(data_births_monthly, formula, predict = F)

@@

```{r}
## Estonia
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_nl_fit <- fitPredictions(dmm_EE_nl, formula)

@@

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_nl_fit <- fitPredictions(data_births_monthly_EE_not_last, formula,
                                predict = F)

@@

```{r}
## TwinR
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_nl_fit <- fitPredictions(dmm_orig_nl, formula)

@@

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_nl_fit <- fitPredictions(data_births_monthly_not_last, formula, predict = F)

@@

```{r}
library(ggplot2)
#some nice colors
bc <- c("purple", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")

@@

### Fig 1: Estonian vs TwinR Full Data

@@

```{r}

fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                 aes(x=births_total, y=prob_twin,
                     color="orig. rural", fill="orig. rural"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian() +
    labs(subtitle = "Model prediction + data mean with SE",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

@@

### Fig 2: Estonian vs Others No Last Birth

@@

```{r}
fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data_nl,
              aes(y = estimate,  x=births_total, color="EE all"),  size = 1) +
    stat_summary(data=dmm_EE_nl_plot,
                 aes(x=births_total, y=prob_twin, color="EE all", fill = "EE all"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data_nl,
                aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
                    color="EE all", fill = "EE all"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data_nl,
             aes(y = estimate,  x=births_total, color="orig. rural"),  size = 1) +
    stat_summary(data=dmm_orig_nl_plot,
                 aes(x=births_total, y=prob_twin,
                     color="orig. rural", fill="orig. rural"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data_nl,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="orig. rural"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18, limits = c(1,NA)) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian(ylim=c(0,0.03)) +
    labs(subtitle = "Without last children, model prediction + data mean with SE",
         y="Per-birth twin. prob.",
         x="Maternal total births")
p3 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "dataset")

@@

>>>>

---
title: 'Suplementary Material 1: Code to Reproduce Analysis and Plots'
author: "Richard Meitern"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!is.null(knitr::opts_knit$get("rmarkdown.pandoc.to"))){
 if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "docx"){
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
} 
}

options(digits = 2)
options(scipen = 9)
options(knitr.kable.NA="")
```

@@

## Data Import Estonia

@@

## Data Import 8 European Populations

@@

```{r}
#compared to twinR original add data on :
#-age at first birth (AFB)
#-quantiles range mother birth year
#-quantiles range offspring birth year
#-mean + SE total births
all_tbls <- rbind(build_data_summary.table(data_births_monthly),
                  build_data_summary.table(data_births_monthly_not_last),
                  build_data_summary.table(data_births_monthly_EE),
                  build_data_summary.table(data_births_monthly_EE_not_last))
```

```{r}
all_tbls$Dataset  <- rep(c("All births", "Without last birth"),2)
all_tbls$Populations  <- rep(c("8 European", "Estonia"),each=2)
all_tbls <- as.data.frame(all_tbls)
row.names(all_tbls) <- paste0(all_tbls$Populations, " populations ",
                              tolower(all_tbls$Dataset)) 
knitr::kable(t(all_tbls),
             caption = paste0("Summary data on the Estonian and other ",
                               "European populations for both ",
                              "all births and wihout last birth for same mothers"))
```

@@

```{r}
# import the function to do model fit and predictions
source("./R/fit_models.R")
```

@@

```{r}
## Estonia - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_fit <- fitPredictions(dmm_EE, formula, predict = T)

@@

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_fit <- fitPredictions(data_births_monthly_EE, formula, nb_boot = 0)

@@

```{r}
## TwinR - mother level data
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_fit <- fitPredictions(dmm_orig, formula, predict=T)

@@

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_fit <- fitPredictions(data_births_monthly, formula, nb_boot = 0)

@@

```{r}
## Estonia
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total"
dmm_EE_nl_fit <- fitPredictions(dmm_EE_nl, formula, predict = T)

@@

```{r}
## Estonia - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id)"
dbm_EE_nl_fit <- fitPredictions(data_births_monthly_EE_not_last, formula,
                                predict = T, nb_boot = 0)

@@

```{r}
## TwinR
formula <- "cbind(twin_total, singleton_total) ~ 1 + births_total + (1|pop)"
dmm_orig_nl_fit <- fitPredictions(dmm_orig_nl, formula, predict = T)

@@

```{r}
## TwinR - birth level data
formula <- "twin ~ 1 + poly(cbind(age, parity), 3) + (1|maternal_id) + (1|pop)"
dbm_orig_nl_fit <- fitPredictions(data_births_monthly_not_last, formula, nb_boot = 0)

@@

```{r}
library(ggplot2)
#some nice colors
bc <- c("azure3","purple4", "black", "navy", "darkgoldenrod2", "springgreen3", "gray")

@@

### Fig 1A: Estonian vs TwinR Full Data

@@

```{r, fig.height=11.7/2, fig.width=8.3}
infoTxt <- "model prediction with 95% CI and data mean SE"
fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data,
              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
    stat_summary(data=dmm_EE[dmm_EE$births_total <17, ],
                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    color="Estonia", fill = "Estonia"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data,
             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
    stat_summary(data=dmm_orig[dmm_orig$births_total <19, ],
                 aes(x=births_total, y=prob_twin,
                     color="8 European", fill="8 European"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="8 European"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian() +
    labs(subtitle = paste0("All births, ", infoTxt),
         y="Per-birth twin. prob.",
         x="Maternal total births")
p2 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "population")

@@

### Fig 1B: Estonian vs Others No Last Birth

@@

```{r, fig.height=11.7/2, fig.width=8.3}
fig2_ext_orig <- ggplot() +  
  geom_line(data=fig2_EE_plot_data_nl,
              aes(y = estimate,  x=births_total, color="Estonia"),  size = 1) +
    stat_summary(data=dmm_EE_nl_plot,
                 aes(x=births_total, y=prob_twin, color="Estonia", fill = "Estonia"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_EE_plot_data_nl,
                aes(y = estimate, x=births_total, ymin = lwr, ymax = upr,
                    color="Estonia", fill = "Estonia"),
                alpha = 0.3) + 
    geom_line(data=fig2_orig_plot_data_nl,
             aes(y = estimate,  x=births_total, color="8 European"),  size = 1) +
    stat_summary(data=dmm_orig_nl_plot,
                 aes(x=births_total, y=prob_twin,
                     color="8 European", fill="8 European"),
                 alpha=0.5,
                 fun.data=mean_se) +
    geom_ribbon(data=fig2_orig_plot_data_nl,
                aes(y = estimate,  x=births_total, ymin = lwr, ymax = upr,
                    fill="8 European"),
                alpha = 0.3) +
    ggplot2::scale_x_continuous(breaks = 1:18, limits = c(1,NA)) + 
    ggplot2::scale_y_continuous(breaks = seq(0,0.03, by=0.005)) +
    ggplot2::coord_cartesian(ylim=c(0,0.03)) +
    labs(subtitle = paste0("Without last birth, ", infoTxt),
         y="Per-birth twin. prob.",
         x="Maternal total births")
p3 <- fig2_ext_orig + base_theme(larger=8) + scale_color_manual(values=bc) + 
  scale_fill_manual(values=bc) + guides(color="none") + labs(fill = "population")

@@

### Figure 2: Model Mean Predictions from Birth Level Data

```{r}
#use new draw_fig_4C function 
source("./R/twinR_theme.R")
```

```{r}
#extract predictions
fig4C_EE_plot_data <- dbm_EE_fit$results
fig4C_EE_plot_data$twin <- fig4C_EE_plot_data$estimates

fig4C_orig_plot_data <- dbm_orig_fit$results
fig4C_orig_plot_data$twin <- fig4C_orig_plot_data$estimates

fig4C_EE_nl_plot_data <- dbm_EE_nl_fit$results
fig4C_EE_nl_plot_data$twin <- fig4C_EE_nl_plot_data$estimates
fig4C_EE_nl_plot_data$parity <- fig4C_EE_nl_plot_data$parity + 1

fig4C_orig_nl_plot_data <- dbm_orig_nl_fit$results
fig4C_orig_nl_plot_data$twin <- fig4C_orig_nl_plot_data$estimates
fig4C_orig_nl_plot_data$parity <- fig4C_orig_nl_plot_data$parity + 1
```

```{r, fig.height=11.7, fig.width=8.3}

rescale_cd <- scale_y_continuous(limits = c(NA, 0.035))

fig2a <- draw_fig_4C(fig4C_orig_plot_data) + 
  labs(subtitle = "(A) 8 European pop., all births")

fig2b <- draw_fig_4C(fig4C_EE_plot_data) + 
  labs(subtitle = "(B) Estonia, all births")

fig2c <- draw_fig_4C(fig4C_orig_nl_plot_data, y_lims = c(0.01, 0.035)) + 
  labs(subtitle = "(C) 8 European pop., without last birth")

fig2d <- draw_fig_4C(fig4C_EE_nl_plot_data,y_lims = c(0.01, 0.035)) + 
  labs(subtitle = "(D) Estonia, without last birth")

gridExtra::grid.arrange(fig2a + guides(), fig2b, fig2c, fig2d, ncol=2) 
```
